# Reproducible Research: Peer Assessment 1

Author: Jody P. Abney

Submission Due Date: 20-Jul-2014 7:30 PM EDT

##Description
This assignment makes use of data from a personal activity monitoring device 
such as a [Fitbit](http://www.fitbit.com/), 
[Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or 
[Jawbone Up](https://jawbone.com/up). This device collects data at 5 minute 
intervals through out the day. The data consists of two months of data from an 
anonymous individual collected during the months of October and November, 2012 
and include the number of steps taken in 5 minute intervals each day.

## Assumptions/Prerequisites
* The basic directory structure begins consists of the following:
    * data - containing source data and analyzed data files
    * doc - containing instructions and finalized documentation
    * figures - containing figures produced from the analysis
    * instructions_fig - containing figures related to the instructions
    * Rcode - containing R code files used in thr analysis
    * Text - containing text, HTML, markdown, and Rmarkdown files

* Required libraries include: 
    * ggplot2
    * lattice
    * knitr

## Loading and preprocessing the data
```{r Load Required Lirbaries, echo = TRUE}
library(ggplot2)
library(lattice)
```

```{r Read Data, echo = TRUE}

## read the activity.csv source data
activity <- read.csv("../data/activity.csv")

```

Execute a "head" command to review the first 10 rows of "activity" data
```{r Head Data, echo = TRUE}

head(activity, n = 10)

```

Execute a "tail" command to get a view of the last 10 rows of "actvity" data
```{r Tail Data, echo = TRUE}

tail(activity, tail = 10)

```

Execute a 'summary' command to get a view of the data elements and any missing
values
```{r Summary Data, echo = TRUE}

summary(activity)

```

Given the number of missing data "NA", care must be exercised to deal with these
missing values in later assignment points.

## What is mean total number of steps taken per day?

### 1. Make a histogram of the total number of steps taken per day

```{r Total Steps Histogram, echo = TRUE}
#### Aggregate the total steps per day data
steps <- with(activity,
              aggregate(steps,
                        by = list(date),
                        sum
              )
) # returns "date" and "total steps per day" data

## Set up the column names
names(steps) <- c("Date", 
                  "Total.Steps")

## Plot the histogram of the total number of steps taken each day
with(steps, 
     hist(Total.Steps, 
          col = "blue", 
          breaks = 30,
          main = "Histogram: Total Steps per Day",
          xlab = "Total Steps",
          ylab = "Frequency (Number of Days)"
     )
)

```

## What is the average daily activity pattern?
### 2. Calculate and report the **mean** and **median** total number of steps taken per day

```{r Mean and Median Total Steps, echo = TRUE}
#### Mean of total steps (must ignore NA values)
steps.mean <- mean(steps$Total.Steps, na.rm = TRUE)
cat("Mean of Total Steps per Day is", steps.mean)

#### Median of total steps (must ignore NA values)
steps.median <- median(steps$Total.Steps, na.rm = TRUE)
cat("Median of Total Steps per Day is", steps.median)

```

## Imputing missing values
* Strategy used for imputing missing values was to use the mean of the 5-minute 
interval that corresponds to the NA value
    1. Make a copy of the activity data without the missing value rows
    2. Compute the "mean" of the steps for each 5-minute interval across
    all of the days
    3. Replace the missing values with the mean for that specific 5-minute
    interval
    4. The dataset with imputed values is named "tidy.activity"

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r Total Number of Missing Value Rows, echo = TRUE}
incomplete <- activity[!(complete.cases(activity)),]
cat("Total number of rows with NAs in the 'activity' dataset:", nrow(incomplete))
```

```{r Imputing Missing Values, echo = TRUE}

# make a copy of activity without the missing data rows
temp <- activity[!(is.na(activity$steps)),]

missing.mean.5 <- with(temp, 
               aggregate(steps, 
                         by = list(interval), 
                         FUN = mean))
names(missing.mean.5) <- c("interval", "Missing.Mean.Interval")

###### Replace the missing values with the missing.mean.5 value
tidy.activity <- activity

for(i in 1:nrow(tidy.activity)) { #
    if(is.na(tidy.activity$steps[i])) {
        temp.5 <- tidy.activity$interval[i]
        
        ## Calculate the lookup row in missing.mean.5
        for(j in 1:nrow(missing.mean.5)) {
            if(temp.5 == missing.mean.5$interval[j]) {
                tidy.activity$steps[i] <- 
                    missing.mean.5$Missing.Mean.Interval[j]
            }
        }
    }
}

```
Verify missing values are now imputed by executing the 'summary' command on the
"tidy.acvitity" dataset to ensure no NA values remain

```{r Summary Imputed Dataset, echo = TRUE}

summary(tidy.activity)

```

#### Make a histogram of the total number of steps taken per day based on the "tidy.activity" data
```{r tidy.activity Histogram, echo = TRUE}

total.tidy.activity <- with(tidy.activity,
                            aggregate(steps,
                                      by = list(date),
                                      sum
                            )
)
names(total.tidy.activity) <- c("Date", "Total.Steps")

with(total.tidy.activity, 
     hist(Total.Steps, 
          col = "blue", 
          breaks = 30,
          main = "Histogram: Total Steps per Day Based on \"tidy.activity\" Dataset",
          xlab = "Total Steps",
          ylab = "Frequency (Number of Days)"
     )
)

```

### Calculate and report the **mean** and **median** total number of steps taken per day based on the "tidy.activity" dataset

```{r Mean and Median Total "tidy.activity" Steps, echo = TRUE}
#### Mean of total steps (must ignore NA values)
steps.mean <- mean(total.tidy.activity$Total.Steps)
cat("Mean of Total Steps per Day based on \"tidy.activity\" dataset is", steps.mean)

#### Median of total steps (must ignore NA values)
steps.median <- median(total.tidy.activity$Total.Steps)
cat("Median of Total Steps per Day based on \"tidy.activity\" dataset is", steps.median)

```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and 
“weekend” indicating whether a given date is a weekday or weekend day.

```{r Add Day.Type Factor, echo = TRUE}

tidy.activity$Weekday <- weekdays(as.Date(tidy.activity$date))
tidy.activity$Day.Type <- ifelse(weekdays(as.Date(tidy.activity$date)) %in% c("Saturday", "Sunday"), 
                                 "weekend", 
                                 "weekday")

```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged across 
all weekday days or weekend days (y-axis).

```{r Panel Plot for Day.Type, echo = TRUE}

temp <- tidy.activity
temp$Weekday <- weekdays(as.Date(temp$date))
temp$Day.Type <- ifelse(weekdays(as.Date(temp$date)) %in% c("Saturday", "Sunday"), "weekend", "weekday")

mean.5.day.type <- with(temp[!is.na(temp$steps),], 
                        aggregate(steps,
                                  by = list(Day.Type, interval),
                                  FUN = mean))
names(mean.5.day.type) <- c("Day.Type",
                            "Interval",
                            "Mean.Steps")

xyplot(Mean.Steps ~ Interval | Day.Type, 
       mean.5.day.type, 
       type = "l", 
       layout = c(1,2))

```

